---
title: "E-GEOD-57452"
author: "ukke"
date: "3 december 2021"
output: html_document
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center')
knitr::opts_chunk$set(out.width = '60%')
knitr::opts_chunk$set(fig.width=12, fig.height=8)


library("mogene10sttranscriptcluster.db")
library("ArrayExpress")
library("arrayQualityMetrics")
library("ggplot2")
#library("huex10sttranscriptcluster.db")
library("limma")
library("oligo")
library("siggenes")
library("affy")
#library("pd.huex.1.0.st.v2")

library("wateRmelon")
```

```{r}
library(affy)
library(arrayQualityMetrics)
library(ArrayExpress)
library(limma)
library(siggenes)
```



#E-GEOD-57452

### General info

The array used for this dataset is A-AFFY-130 - Affymetrix GeneChip Mouse Gene 1.0 ST Array [MoGene-1_0-st-v1]. Mice were infected with influenza and RNA was extracted from the lungs after 10 days. We used samples involving susceptible mice after 10 days of infection with influenza from this dataset.

### Intensity values

Read in the microarray data and display the head and dimensions of the intensity value matrix.
```{r, message=F}
id_1 <- "E-GEOD-57452"
exonCELs <- list.celfiles("../Datasets/Microarray1")
data.raw_1 <- read.celfiles(paste(rep("../Datasets/Microarray1/",length(exonCELs)),exonCELs,sep=""))
```


```{r}
# make vector containing the sample class
samples <- c(replicate(3, "control"), "day0", "day1", "day2", replicate(2, "day3"), replicate(2, "day4"), replicate(3, "day5"), "day6", replicate(2, "day7"), replicate(2, "day8"), replicate(2, "day9"), replicate(3, "day10")) 

# add the sample classes to the pData object
pData(data.raw_1)[,2] <- samples 

colnames(pData(data.raw_1)) <- c("index", "treatment")


# filter control samples and samples taken after 3 days of infection
filter <- colnames(data.raw_1)[data.raw_1@phenoData@data$treatment=="control" | data.raw_1@phenoData@data$treatment=="day10"]

# apply filter
filtered <- data.raw_1[,filter]

# check dimentsions of filtered object
dim(exprs(filtered))
```


```{r}
## arrayQualityMetrics 
arrayQualityMetrics(filtered,outdir="../Datasets/microarray1/raw",force=T)
arrayQualityMetrics(filtered,outdir="../Datasets/microarray1/rawlog",force=T,do.logtransform=T)

# Preprocessing (using the oligo function because affy didnt work)
MouseRMA<- oligo::rma(filtered,background=T)

## QC post preprocessing
arrayQualityMetrics(MouseRMA,outdir="../Datasets/microarray1/rma",force=T)  			#RMA produces log-transformed data


```

##Data exploration

```{r}
# transpose the data before PcA as this function requires the variables to b columns
data <- t(as.data.frame(MouseRMA@assayData$exprs))
pca <- prcomp(data, center = T, scale. = T)

summary(pca)
# save as dataframe and add treatment variable
pca_out <- as.data.frame(pca$x)
pca_out$treatment <- as.character(MouseRMA@phenoData@data$treatment)
pca_out$sample <- c("rep1", "rep2", "rep3", "rep1", "rep2", "rep3")

# get lablels
percentage <- round(pca$sdev / sum(pca$sdev) * 100, 2)
percentage <- paste( colnames(pca_out), "(", paste( as.character(percentage), "%", ")", sep="") )

ggplot(data = pca_out)+
  geom_point(aes(x = PC1, y = PC2, colour = treatment, label=sample), size=3)+
  geom_text(aes(x = PC1, y = PC2, colour = treatment, label=sample),hjust=0.5, vjust=1.15)+
  theme_bw()+
  xlab(percentage[1])+
  ylab(percentage[2])+
  labs(colour = "treatment")+
  theme(plot.title = element_text(hjust = 0.5))+
  scale_colour_manual(values = c("firebrick", "midnightblue"))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

ggsave("PCA_array1.png", dpi=750, width=8, height = 5)
```



determine differential expression

```{r}
annot <- factor(pData(MouseRMA)[,2])

## Differential expression by LIMMA
# Method as stated in limma package (no intercept, easy for simple model designs)
design <- model.matrix(~0+annot)
colnames(design)<-c("control","infected")

# make linear model
fit <- lmFit(MouseRMA,design)

# create contrast matrix to get the differential expression between samples from infected mice and uninfected mice 
cont.matrix <- makeContrasts(NvsS=control-infected,levels=design)
fit2 <- contrasts.fit(fit,cont.matrix) 
fit2 <- eBayes(fit2)

# make MA plot for model with applied contrast matrix
limma::plotMA(fit2)
```

```{r}
library(ggplot2)
# DE results with multiple testing correction (Benjamini-Hochberg = BH)
LIMMAout <- topTable(fit2,adjust="BH",number=nrow(exprs(MouseRMA)))



# add column indicating for all differntially expressed genes (adjusted p-value < 0.05) whether they're up- or down-regulated

LIMMAout$diffexpressed <- "NO"
LIMMAout$diffexpressed[LIMMAout$logFC > 0 & LIMMAout$adj.P.Val < 0.05] <- "UP"
LIMMAout$diffexpressed[LIMMAout$logFC < 0 & LIMMAout$adj.P.Val < 0.05] <- "DOWN"

# do the same but if we would not correct for multiple testing.
LIMMAout$diffexpressed_no_BH <- "NO"
LIMMAout$diffexpressed_no_BH[LIMMAout$logFC > 0 & LIMMAout$P.Value < 0.05] <- "UP"
LIMMAout$diffexpressed_no_BH[LIMMAout$logFC < 0 & LIMMAout$P.Value < 0.05] <- "DOWN"

# code to make volcano plots

ggplot(data = LIMMAout, aes(x= logFC, y = -log10(adj.P.Val), colour = diffexpressed)) +
 geom_point()+
 theme_bw()+
  geom_hline(yintercept = -log10(0.05), linetype="dashed", color="midnightblue")+
  annotate("text", min(4), 1.3, vjust = -1, label = "Cutoff", color="midnightblue")+
 labs(colour = "Differentialy expressed")

ggsave("volcanoplot.png", dpi=750)

ggplot(data = LIMMAout, aes(x= logFC, y = -log10(P.Value), colour = diffexpressed_no_BH)) +
 geom_point()+
 theme_bw()+
  geom_hline(yintercept = -log10(0.05), linetype="dashed", color="midnightblue")+
  annotate("text", min(4), 1.3, vjust = -1, label = "Cutoff", color="midnightblue")+
 labs(colour = "Differentialy expressed")

table(LIMMAout$diffexpressed)

```


### Annotation

```{r}
# get the annotation trough package (mogene10sttranscriptcluster.db) found at https://www.biostars.org/p/9477569/
columns(mogene10sttranscriptcluster.db)

annotTable <- select(
  mogene10sttranscriptcluster.db,
  keys = keys(mogene10sttranscriptcluster.db),
  column = c('PROBEID', 'SYMBOL', 'ENTREZID', 'ENSEMBL', 'GENENAME', 'PROSITE'),
  keytype = 'PROBEID')


## sort annotation data alphabetically on probe name

annotTable.filt <- annotTable[sort(annotTable$PROBEID,index.return=T)$ix,]

# merge information from multiple lines describing the same probe
probe <- "start"
position <- 0
for (i in 1:dim(annotTable.filt)[1]){
  if (annotTable.filt[i, 1] != probe){
    probe <- annotTable.filt[i, 1]
    position <- i
  }
  else{
    # concatenate the information o f the 2 lines with a ; as separator
    annotTable.filt[position,2:5] <- paste(annotTable.filt[position,2:5], annotTable.filt[i, 2:5], sep="; ")
    # mark the line 
    annotTable.filt[i,1] <- NA
  }
}

annotTable.filt <- annotTable.filt[!is.na(annotTable.filt$PROBEID),]

## Check if all probes are present in both sets
dim(annotTable.filt)
dim(LIMMAout)

 ## Double check => "Assumption is the mother of all fuck up's ;)"
sum(annotTable.filt$PROBEID!=sort(rownames(LIMMAout)))

## Sort LIMMA output alphabetically on probe name
LIMMAout_sorted <- LIMMAout[sort(rownames(LIMMAout),index.return=T)$ix,]

## Add gene names to LIMMA output
LIMMAout_sorted$gene <- annotTable.filt$SYMBOL
LIMMAout_annot <- LIMMAout_sorted[sort(LIMMAout_sorted$adj.P.Val,index.return=T)$ix,]
LIMMAout_annot[LIMMAout_annot$diffexpressed != "NO",]
LIMMAout_annot[LIMMAout_annot$diffexpressed != "NO",8]

# determine how many differentially expressed probes have an annotated gene
table(is.na(LIMMAout_annot[LIMMAout_annot$diffexpressed != "NO",9]))


## alternative annotation method:

# annotation file from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL6246
annotation_MA <- read.delim("C:/Users/tobia/Downloads/GPL6246.annot",header=T, sep="\t", skip=27, fill=T)
print(head(annotation_MA))
annotation_MA <- annotation_MA[sort(annotation_MA$ID,index.return=T)$ix,]

dim(annotation_MA)
dim(LIMMAout)



## the dimensions don't match but we can still check whether the information for the differentially expressed probes matches using both annotation methods

annotation_MA[annotation_MA$ID %in% rownames(LIMMAout_annot[LIMMAout_annot$diffexpressed != "NO",]),]


```